**The Heart can Lie: A Preliminary Investigation of the Role of Interoception and Theory of Mind in Deception**

### 2024-09-15 paper/manuscript.Rmd
minor
R3

Although few studies exist that investigate the relationship between interoception and deceptive ability *per se*, previous decision-making studies have demonstrated a negative correlation between interoceptive awareness and one's likelihood to make risky decisions [@dunn2010listening; @furman2013interoceptive]. This is in line with the somatic marker hypothesis, which posits that an accurate evaluation of one's bodily signals facilitates the use of such interoceptive feedback to guide rational decision making [@damasio1996somatic]. Indeed, @sugawara2020effects further reported that individuals who received interoceptive training were more likely to show higher interoceptive accuracy and make reasoned decisions. Given that deciding to lie generally involves a consideration of the potential costs of getting caught, and hence could also be perceived as risky behavior [@kireev2013possible], interoception could be construed to be negatively related to lying ability. However, some studies have instead found heightened interoceptive attention (one's self-focus towards internal bodily signals), to predict immoral behaviour, such as cheating [@ditto2006visceral; @williams2016valjean; @lenggenhager2013listening]. Extending these findings to social cognition, @vabba2022interoceptive further reports individuals with lower interoception told significantly less egoistic lies when the social reputational stakes were high, whereas individuals with higher interoception did not exhibit a significant difference in the number of lies told. Given the scarce research on interoception and deception, more studies are herein needed to clarify these mixed findings.

@@

Thirty university students from Singapore were recruited through posters, flyers, and online social media platforms, and rewarded with study credits for their time.
<!-- Inclusion criteria included residing in Singapore, and an absence of current or history of neurological, psychiatric, or cardiovascular disorders.  -->
Four participants were excluded as their data was not recorded due to technical issues. The final sample consists of 26 participants (Mean age = 20.9, SD = 2.0, range:[18, 25], Sex: 65.4% women, 34.6% men). The heart rate of one participant and response time of one participant were excluded from further analysis due to extreme outlying values. To maximize statistical power, the problematic data from these 2 participants were only excluded from analyses involving those measures; all other data were retained for analyses.

@@

Two measures of ToM and its related constructs were administered. The Yoni Task [@shamay2007dissociable] is a behavioral task which assesses first and second-order ToM abilities in both cognitive and affective domains. Participants were presented with the face of a character named "Yoni", surrounded by four colored pictures of objects or faces, one in each corner of the screen. In total, each participant completed 101 trials - 49 trials assessing their affective ToM abilities, 37 trials assessing their cognitive ToM abilities and 15 control trials. During each trial, participants were instructed to respond based on specific corresponding cues such as the directions of Yoni's eye gaze, facial expressions etc., In the control trials, participants made judgements based on Yoni's physical context (physical ToM). More specifically, in first-order trials, participants were instructed to make inferences about Yoni's mental state with regards to the objects surrounding it (e.g., "Yoni is thinking of..." for cognitive ToM trials or "Yoni likes..." for affective ToM trials). In more complex second-order trials, participants had to correctly infer the interaction between Yoni and others' mental states (e.g., "Yoni is thinking of the fruit that ... wants" for cognitive ToM trials or "Yoni likes the fruit that ... likes" for affective ToM trials).

@@

Using PsychoPy [@peirce2019psychopy2], we implemented a directed-lying task in which participants were instructed to briefly answer 80 questions [taken from their previously taken Autobiographical Memory Questionnaire - AMQ, @rubin2003belief] pertaining to their personal preferences and subjective experiences, by either lying or telling the truth (depending on whether they see "lie" or "truth" written on the screen). Their goal was to make convincing answers, so that truths would be judged as truths by the receiver, and lies as lies. The nature of the receiver was different depending on the condition: participants were told that for half of the trials, they would have to convince another participant that would be observing them from a separate room (COVID regulations were used as a justification) via a webcam connection (*Interrogation* condition). For the remaining trials, participants were tasked to convince a "lie detection machine" that would be assessing their behaviour through their physiological signals (*Polygraph* condition). In reality, there was no real "receiver" and their answers were not judged externally (the study focused on their subjective ratings and reactions). The two conditions were presented in a counter-balanced order, and each comprised of 40 trials (20 truth; 20 lies).

@@

Finally, on top of the deception task, we also measured participants' self-reported tendency to lie in their everyday life using the *Lie Scale* [@makowski2021structure], a 16-item questionnaire that assesses 4 dispositional lying dimensions - Ability ($\alpha = 0.92$), Frequency ($\alpha = 0.66$), Negativity ($\alpha = 0.66$), and Contextuality ($\alpha = 0.70$).

@@

A within-subjects design was used in the present study, which is comprised of 2 sessions, to investigate the roles interoception and ToM play in lying ability. During session 1, participants answered a brief demographic survey as well as a questionnaire regarding their personal preferences and subjective experiences (the AMQ), followed by a series of psychological scales (i.e., BES, MAIA and Lie scale), which were randomly displayed.

@@

To maximize the signal-to-noise ratio, we performed a feature reduction on our two groups of predictor variables (namely, ToM and interoception) using factor analysis over PCA, as the goal was to extract meaningful and consistent factors, rather than merely maximizing the variance explained. Then, we modelled the relationship between these inter-individual composite scores (note that the analysis for all individual variables is nonetheless included in the analysis report) and the three outcome variables in interaction with the condition (polygraph *vs.* interrogation). Finally, we investigated the relationship between the deception scale traits, and the ToM and interoception scores using Bayesian correlations. As all the analyses and data have been made available, we will in the manuscript focus on significant, i.e., - in this context - statistically reliable and in our opinion theoretically relevant.

@@

Compared to truths, lies were rated with less confidence ($\Delta=-1.35,~95\%~CI~[-1.46, -1.23],~pd = 100\%$), but no significant difference between the conditions was found. On the other hand, the RT did not differ between truths and lies, but was significantly slower in the polygraph condition for both conditions ($\Delta=0.25$, 95\% CI [0.62, 0.41], pd = 100\%). The heart rate was significantly more elevated during lies as compared to truths ($\Delta=1.16$, 95\% CI [0.57, 1.73], pd = 100\%), and during interrogation as compared to the polygraph condition ($\Delta=4.84$, 95\% CI [4.23, 5.44], pd = 100\%).

The indirect phrasing of the question only had a significant effect on RT ($\beta=0.36$, 95\% CI [0.21, 0.51], pd = 100\%), leading to slower answers, regardless of whether they were lies or truths. Given this absence of interaction with the type of answers in any modality, this factor was not included in subsequent analysis.

@@

The higher composite *ToM* score was significantly associated with a decreased confidence in lies ($\beta=-0.19$, 95\% CI [-0.36, -0.02], pd = 98.47\%), specifically in the polygraph condition. **Figure 1** illustrates the interindividual correlates of lying confidence. The higher composite *ToM* score was also associated with slower answers for lies ($\beta=0.42$, 95\% CI [0.01, 0.83], pd = 97.67\%), specifically in the polygraph condition. No significant effect was found with regards to dispositional lying traits, heart rate, and RT for truths in both polygraph and interrogation conditions.

>>>>

Although few studies exist that investigate the relationship between interoception and deceptive ability *per se*, previous decision-making studies have demonstrated a negative correlation between interoceptive awareness and one's likelihood to make risky decisions [@dunn2010listening; @furman2013interoceptive]. This is in line with the somatic marker hypothesis, which posits that an accurate evaluation of one's bodily signals facilitates the use of such interoceptive feedback to guide rational decision making [@damasio1996somatic]. Indeed, @sugawara2020effects further reported that individuals who received interoceptive training were more likely to show higher interoceptive accuracy and make reasoned decisions. Given that deciding to lie generally involves a consideration of the potential costs of getting caught, and hence could also be perceived as risky behavior [@kireev2013possible], interoception could be construed to be negatively related to lying ability. However, some studies have instead found heightened interoceptive attention (one's self-focus towards internal bodily signals), to predict immoral behaviour, such as cheating [@ditto2006visceral; @williams2016valjean; @lenggenhager2013listening]. Extending these findings to social cognition, @vabba2022interoceptive further reports individuals with lower interoception told significantly fewer egoistic lies when the social reputational stakes were high, whereas individuals with higher interoception did not exhibit a significant difference in the number of lies told. Given the scarce research on interoception and deception, more studies are herein needed to clarify these mixed findings.

@@

Thirty university students from Singapore were recruited through posters, flyers, and online social media platforms, and rewarded with study credits for their time.
<!-- Inclusion criteria included residing in Singapore, and an absence of current or history of neurological, psychiatric, or cardiovascular disorders.  -->
Four participants were excluded as their data was not recorded due to technical issues. The final sample consisted of 26 participants (Mean age = 20.9, SD = 2.0, range:[18, 25], Sex: 65.4% women, 34.6% men). The heart rate of one participant and response time of one participant were excluded from further analysis due to extreme outlying values. To maximize statistical power, the problematic data from these two participants were only excluded from analyses involving those measures; all other data were retained for analyses.

@@

Two measures of ToM and its related constructs were administered. The Yoni Task [@shamay2007dissociable] is a behavioral task which assesses first and second-order ToM abilities in both cognitive and affective domains. Participants were presented with the face of a character named "Yoni", surrounded by four colored pictures of objects or faces - one in each corner of the screen. In total, each participant completed 101 trials - 49 trials assessing their affective ToM abilities, 37 trials assessing their cognitive ToM abilities and 15 control trials. During each trial, participants were instructed to respond based on specific corresponding cues such as the directions of Yoni's eye gaze, facial expressions etc., In the control trials, participants made judgements based on Yoni's physical context (physical ToM). More specifically, in first-order trials, participants were instructed to make inferences about Yoni's mental state with regards to the objects surrounding it (e.g., "Yoni is thinking of..." for cognitive ToM trials or "Yoni likes..." for affective ToM trials). In more complex second-order trials, participants had to correctly infer the interaction between Yoni and others' mental states (e.g., "Yoni is thinking of the fruit that ... wants" for cognitive ToM trials or "Yoni likes the fruit that ... likes" for affective ToM trials).

@@

Using PsychoPy [@peirce2019psychopy2], we implemented a directed-lying task in which participants were instructed to briefly answer 80 questions [taken from their previously taken Autobiographical Memory Questionnaire - AMQ, @rubin2003belief] pertaining to their personal preferences and subjective experiences, by either lying or telling the truth (depending on whether they see "lie" or "truth" written on the screen). Their goal was to make convincing answers, so that truths would be judged as truths by the receiver, and lies as lies. The nature of the receiver was different depending on the condition: participants were told that for half of the trials, they would have to convince another participant who would be observing them from a separate room (COVID regulations were used as a justification) via a webcam connection (*Interrogation* condition). For the remaining trials, participants were tasked to convince a "lie detection machine" that would be assessing their behaviour through their physiological signals (*Polygraph* condition). In reality, there was no real "receiver" and their answers were not judged externally (the study focused on their subjective ratings and reactions). The two conditions were presented in a counter-balanced order, and each comprised of 40 trials (20 truth; 20 lies).

@@

Finally, on top of the deception task, we also measured participants' self-reported tendency to lie in their everyday life using the *Lie Scale* [@makowski2021structure], a 16-item questionnaire that assesses four dispositional lying dimensions - Ability ($\alpha = 0.92$), Frequency ($\alpha = 0.66$), Negativity ($\alpha = 0.66$), and Contextuality ($\alpha = 0.70$).

@@

A within-subjects design was used in the present study, which is comprised of two sessions, to investigate the roles interoception and ToM play in lying ability. During session 1, participants answered a brief demographic survey as well as a questionnaire regarding their personal preferences and subjective experiences (the AMQ), followed by a series of psychological scales (i.e., BES, MAIA and Lie scale), which were randomly displayed.

@@

To maximize the signal-to-noise ratio, we performed a feature reduction on our two groups of predictor variables (namely, ToM and interoception) using factor analysis over PCA, as the goal was to extract meaningful and consistent factors, rather than merely maximizing the variance explained. Then, we modelled the relationship between these inter-individual composite scores (note that the analysis for all individual variables is nonetheless included in the analysis report) and the three outcome variables in interaction with the condition (polygraph *vs.* interrogation). Finally, we investigated the relationship between the deception scale traits, and the ToM and interoception scores using Bayesian correlations. All analyses and data have been made publicly available. Therefore, in this manuscript, we will focus on discussing significant findings, which - in this context - are statistically reliable and in our opinion theoretically relevant results.

@@

Compared to truths, lies were rated with less confidence ($\Delta=-1.35,~95\%~CI~[-1.46, -1.23],~pd = 100\%$), but no significant difference between the conditions was found. On the other hand, the RT did not differ between truths and lies, but was significantly slower in the polygraph condition for both conditions ($\Delta=0.25,~95\% CI [0.62, 0.41],~pd = 100\%$). The heart rate was significantly more elevated during lies as compared to truths ($\Delta=1.16,~95\% CI [0.57, 1.73],~pd = 100\%$), and during interrogation as compared to the polygraph condition ($\Delta=4.84,~95\% CI [4.23, 5.44],~pd = 100\%$).

The indirect phrasing of the question only had a significant effect on RT ($\beta=0.36,~95\% CI [0.21, 0.51],~pd = 100\%$), leading to slower answers, regardless of whether they were lies or truths. Given this absence of interaction with the type of answers in any modality, this factor was not included in subsequent analysis.

@@

The higher composite *ToM* score was significantly associated with a decreased confidence in lies ($\beta=-0.19,~95\% CI [-0.36, -0.02],~pd = 98.47\%$), specifically in the polygraph condition. **Figure 1** illustrates the interindividual correlates of lying confidence. The higher composite *ToM* score was also associated with slower answers for lies ($\beta=0.42,~95\% CI [0.01, 0.83],~pd = 97.67\%$), specifically in the polygraph condition. No significant effect was found with regards to dispositional lying traits, heart rate, and RT for truths in both polygraph and interrogation conditions.
